---
title: "NGS barcoding"
author: "Lambda Moses"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Array}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here I explore the history and geography of spatial transcriptomics based on array capture through the metadata I collected.
```{r}
library(museumst)
library(ggplot2)
library(purrr)
library(dplyr)
library(sf)
theme_set(theme_bw())
```

# Import data
By default, the `read_metadata` function reads the appropriate sheet from a cache within this package. This may be outdated, since the Google Sheets are constantly updated. Use the argument `update = TRUE` to download the most up to date version. The default is `FALSE` to avoid API limits during automated CRAN checks.
```{r}
sheet <- read_metadata("NGS barcoding")
```

# Number of publications
How many publications per year?
```{r}
# Remove duplicates as different datasets in the same publication have their own rows
publications <- get_pubs_df(sheet, c("repo", "accession"))
```

How many publications are there in total in this sheet?
```{r}
nrow(publications)
```

```{r}
# For maps later in this notebook
city_gc <- geocode_inst_city(publications)
pubs_on_map2 <- partial(pubs_on_map, city_gc = city_gc)
```

## Overall
Number of publications per year. Preprints are excluded from plots over years as it takes months to publish so the dates of the preprints are incoherent with those for the other papers. 
```{r}
events <- read_major_events()
```

```{r, fig.width=7, fig.height=3}
events %>% 
  filter(category == "array") %>% 
  plot_timeline(rep(c(1, -1), length.out = nrow(.)), 
                expand_x = c(0.1, 0.1), expand_y = c(0.05, 0.05))
```

By default, preprints are excluded from plots of number of publications over time, since the timing of preprints is incoherent to that of published papers. Including preprints will inflate the number of publications in recent months, and if we make the same plot a few months later, that inflation we see now will be gone and moved to a more recent date. You can add the `preprints = TRUE` argument to all functions plotting things over time in this package to include preprints. You can also use the `binwidth` argument to specify bin width in days. The default is 365 days.
```{r}
pubs_per_year(publications, binwidth = 100, preprints = TRUE)
```

Again, preprints are not included here since their timing is different from that of publications.

## By method
How about when broken down by method (not for data analysis sheets)?

```{r}
methods <- unnest_cat(sheet, method)
```

```{r, fig.width=4, fig.height=4}
pubs_per_year(methods, facet_by = "method", binwidth = 100, preprints = TRUE,
              sort_by = "recent")
```

I don't know how much this actually says due to the small sample size. This is also kind of confusing since Geo-seq really is a form of LCM coupled with SMART-seq, so it should be categorized as LCM, but the authors gave it a new name. 

## By species

```{r}
species <- unnest_cat(sheet, species)
```

```{r}
pubs_per_cat(species, species)
```

```{r, fig.width=4, fig.height=2}
pubs_per_cat(species, species, isotype = TRUE, img_unit = 2)
```

I debated for a while which image to use for humans. In order to avoid racist and sexist connotations, I picked the skull as non-experts can't tell the race and sex from the skull. If you think it's bad to use a skull, then I'll use a Black, Hispanic, Middle Eastern, North African, Central Asian, or South Asian woman's portrait. I stand with the oppressed.

## By journal
```{r}
sort(table(publications$journal))
```

## Location
### General
Just some barplots for number of publications per institution, city, and country. 
```{r}
pubs_per_cat(publications, country)
```

How about per capita? Actually this probably does not say much since many researchers in the West are immigrants. Then it raises the burning political question of who counts as the "population".

```{r}
pubs_per_capita(publications, plot = "bar")
```

How about country over time? There might not be enough publications to show a trend.
```{r}
pubs_per_year(publications, facet_by = "country", binwidth = 150)
```
Now look at cities. 

```{r}
pubs_per_cat(publications, city)
```

Institutions
```{r}
pubs_per_cat(publications, institution)
```

OK, now here comes the maps!

### Worldwide
```{r}
pubs_on_map2(publications)
```

Let me also plot the per capita thing on a map, as a choropleth
```{r}
pubs_per_capita(publications)
```

Break down by species
```{r, fig.width=8, fig.height=3}
pubs_on_map2(species, facet_by = "species")
```

By method
```{r, fig.width=8, fig.height=4}
pubs_on_map2(methods, facet_by = "method")
```

It seems that ST is the only method that did spread. I think Visium did too; it's just that the papers haven't been published yet.

### Europe
Plot a map just for Europe. A problem here is that if I simply filter the `world` sf object, I'll get some islands away from what we usually think is Europe. Kind of feel bad for Russians since Russia is so large that it makes the map look bad and I don't have a paper from Russia in this spreadsheet. 

```{r}
pubs_on_map2(publications, zoom = "europe")
```

```{r}
pubs_per_capita(publications, "europe")
```

```{r}
pubs_on_map2(species, zoom = "europe", facet_by = "species")
```

Maybe I just made this plot for fun. Not sure what to say about it.

```{r}
pubs_on_map2(methods, zoom = "europe", facet_by = "method")
```

So all the other methods are American

### USA
Also a plot just for America. I did not encounter a publication in this spreadsheet that is from Hawaii or Alaska, but I'll include Hawaii and Alaska in the map anyway in case I get one in the future.

```{r}
pubs_on_map2(publications, zoom = "usa")
```

```{r}
pubs_per_capita(publications, "usa")
```

```{r}
pubs_on_map2(species, zoom = "usa", facet_by = "species")
```

```{r}
pubs_on_map2(methods, zoom = "usa", facet_by = "method")
```

# Word cloud
## Titles
```{r, fig.height=6, fig.width=6}
plot_wordcloud(sheet)
```

```{r, fig.height=5, fig.width=5}
plot_wordcloud(sheet, species_use = "Mus musculus")
```

```{r, fig.height=7, fig.width=7}
plot_wordcloud(sheet, species_use = "Homo sapiens")
```

I think the prominence of "spatial" is probably driven by ST.

## Tissues
```{r, fig.height=6, fig.width=6}
plot_wordcloud(sheet, col_use = "tissue", scale = c(5, 0.1))
```

```{r}
plot_wordcloud(sheet, species_use = "Mus musculus", col_use = "tissue", scale = c(5, 0.1))
```

```{r, fig.height=5, fig.width=5}
plot_wordcloud(sheet, species_use = "Homo sapiens", col_use = "tissue", scale = c(5, 0.1))
```

## Department names

```{r, fig.height=6, fig.width=6}
plot_wordcloud(sheet, col_use = "department", other_stop_words = inst_words, scale = c(5, 0.1))
```

I think that's driven by Department of Gene Technology from Sweden

## Downstream analyses
```{r, fig.height=5, fig.width=5}
plot_wordcloud(sheet, "downstream")
```

# Programming languages
```{r}
langs <- unnest_cat(sheet, language)
```

```{r}
pubs_per_cat(langs, language, isotype = TRUE, img_unit = 2)
```

# Data and code availability

How many publications have provided a code repo? As with any plot over time, preprints are excluded.
```{r}
hist_bool(publications, !is.na(repo), binwidth = 100, preprints = TRUE)
```

How many publications have provided an accession for sequencing data?
```{r}
hist_bool(publications, !is.na(accession), binwidth = 100, preprints = TRUE)
```

```{r}
hist_bool_line(publications, !is.na(accession), binwidth = 100, preprints = TRUE)
```

I don't think it's significant.
```{r}
sessionInfo()
```

